A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-05-27 discussion.tex
rework most useful patterns and add more todos

In this work, we used Ruxanne to successfully mine 20 cross-project bug fix patterns. We presented these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect five of the BC-related patterns. Fundamentally, Clippy cannot detect some of the remaining patterns. For instance, we proposed the pattern \textit{Adding mutability}. In the Rust context, this change anticipates a future change to the code which will perform a mutation of a previously immutable value. Predicting the future is beyond Clippy's remit. To the best of our knowledge, the rest of the patterns (3 BC-related patterns and 12 general patterns) cannot be reported by the current linting tools.

@@

\subsection{Patterns Present and Missing}

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns.

The two most common patterns were general patterns modifying struct fields and attributes, with 179 and 172 instances respectively. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes. Attribute modifications are far more interesting, and account for the second- and fifth-most popular pattern. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and implies that program repair must understand attributes. A further challenge is that different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

The most common borrow-checker-related pattern and third-most-common pattern overall is dropping \texttt{clone()}. This pattern is performance-related; when correctly applied, it has no semantic impact on the program, but reduces resource consumption. Rust is used in situations where performance is important; the number of instances of this pattern confirms that developers do work on performance issues. This implies that program repair tools would benefit from being performance-aware, rather than limiting themselves strictly to functional properties of the code.

@@

The seventh-most-popular pattern is the Rust-specific change of dropping mutability (i.e. making data immutable). Tools like ReImInfer by~\cite{HuangMDE2012} (for a Java extension), and techniques like the one proposed in~\cite{EyolfsonAbstractImmutability} for C++, aim to infer that data is immutable. In the Rust context, it is easier to make data immutable: the semantics for immutability are clear. Immutability is part of the core Rust language. mmutability escape hatches in C++ complicate issues, but in Rust, the escape hatch is the clearly-marked \texttt{unsafe} block. This pattern thus suggests that tools for inferring Rust immutability could be viable.

@@

Conversely, we were surprised to find that Ruxanne reported no patterns associated with certain Rust-related features. For instance, we did not observe any patterns associated with the Rust \verb+move+ operator, which has been known to trouble many beginning Rust programmers. \verb+move+ is used to transfer the ownership of values to a closure. In some cases (e.g. multithreading), not moving the ownership could create invalid references. Rust compiler checks prevent such errors. Our speculation is that, since Rust enforces strict compiler checks, programmers usually get many compile-time errors, but they only push their code once it is free of compile errors. We thus would not observe patterns which do not make it to the repository.

@@

We believe our results can help with the development of IDE and program repair tools for Rust. We have discussed the actionability of some of the general patterns and three BC-related patterns that are not detected by Clippy. For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size. Similarly, we presented the pattern \textit{Dropping clone and adding borrowing}. As discussed, Clippy does not detect this pattern, and repetitive clones can be computationally expensive. Using state-of-the-art methods for program repair, one could design a tool to recognize this pattern and change variable cloning to simple borrowing. These findings and insights can greatly aid researchers in creating effective IDE tools tailored for Rust development. Moreover, the empirical frequency numbers associated with each pattern, highlighting its prevalence, can provide valuable guidance for the development of program repair tools.

A general approach for code embedding is to extract paths with semantic information that better serve a given embedding goal. For instance, to design an embedding method for repairing Rust programs, the designer might want to disregard certain elements, such as moving ownership to closure bodies. Such elements may be irrelevant for repair because they did not appear in the bug patterns that we presented.

@@

In our work there are two main threats to internal validity: (1) A threat to internal validity is confounding, where changes to what shows up in the embedding are not due to changes in the code being embedded. In this case, the weighting scheme may contribute to confounding because we manually adjusted the weights. Nevertheless, our weighting scheme can be readjusted to find different patterns, which is why we made our pipeline publicly available. (2) Our code embedding approach is based on the frequency of observed program elements in ASTs, and we use DBSCAN as our clustering algorithm. Other code embedding methods and clustering algorithms (e.g. SLINK) might output new clusters that our pipeline is unable to find. A threat to external validity is selection bias, where the benchmarks differ systematically from the actual world. The selected projects might cause selection bias and therefore we might not have presented patterns that may exist in other projects. Another threat to external validity can be the developers reporting the commits as bug fixing commits while they are not really fixing a functionality. Similarly, a commit message might not contain our target keywords while the commit is associated with bug fixing changes.

>>>>

In this work, we used Ruxanne to successfully mine 20 cross-project bug fix patterns. We presented these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect 5 of the 8 BC-related patterns. Fundamentally, Clippy cannot detect some of the remaining patterns. For instance, we proposed pattern \textit{Adding mutability}. In the Rust context, this change anticipates a future change to the code which will perform a mutation of a previously immutable value. Predicting the future is beyond Clippy's remit. To the best of our knowledge, the remaining patterns (3 BC-related patterns and 12 general patterns) cannot be reported by the current linting tools.

@@

\subsection{Patterns Present and Missing}
\label{subsec:patterns}

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns. In summary, program repair tools must understand semantics of attributes and should be performance-aware. It may also be useful for tools to be able to automatically infer immutability.

The two most common patterns were general patterns modifying struct fields and attributes, with 179 and 172 instances respectively. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes. Attribute modifications are far more interesting, and account for the second- and fifth-most popular patterns. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and \emph{implies that program repair must understand the effects of attributes}. Yet, different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

The most common borrow-checker-related pattern and third-most-common pattern overall is dropping \texttt{clone()}. This pattern is performance-related; when correctly applied, it has no semantic impact on the program, but reduces resource consumption. Rust is used in situations where performance is important; the number of instances of this pattern confirms that developers do work on performance issues. This implies that \emph{program repair tools would benefit from being performance-aware}, rather than limiting themselves strictly to functional properties of the code.

@@

The seventh-most-popular pattern is the Rust-specific change of dropping mutability (i.e. making data immutable). Tools like ReImInfer by~\cite{HuangMDE2012} (for a Java extension), and techniques like the one proposed in~\cite{EyolfsonAbstractImmutability} for C++, aim to infer that data is immutable. In the Rust context, it is easier to make data immutable: the semantics for immutability are clear. Immutability is part of the core Rust language. Immutability escape hatches in C++ complicate issues, but in Rust, the escape hatch is the clearly-marked \texttt{unsafe} block. This pattern thus suggests that \emph{tools for inferring Rust immutability could be viable}.

@@

Conversely, we were surprised to find that Ruxanne reported no patterns associated with certain Rust-related features. For instance, we did not observe any patterns associated with the Rust \verb+move+ operator, which has been known to plague many beginning Rust programmers. \verb+move+ is used to transfer the ownership of values to a closure. In some cases (e.g. multithreading), not moving the ownership could create invalid references. Rust compiler checks prevent such errors. Our speculation is that, since Rust enforces strict compiler checks, programmers usually get many compile-time errors, but they only push their code once it is free of compile errors. We thus would not observe patterns which do not make it to the repository.

@@

Having discussed the most frequent patterns and their implications on repair tools, we continue by discussing potential features for repair and refactoring tools in more detail. We believe our results can help with the development of program repair and refactoring tools for Rust. We have discussed the actionability of some of the general patterns and three BC-related patterns that are not detected by Clippy.

For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size.

Similarly, we presented the pattern \textit{Dropping clone and adding borrowing}. As discussed, Clippy does not detect this pattern, and repetitive clones can be computationally expensive. Using state-of-the-art methods for program repair, one could design a tool to recognize this pattern and change variable cloning to simple borrowing.

We believe that our insights can greatly aid researchers in creating effective IDE tools tailored for Rust development. Moreover, the empirical frequency numbers associated with each pattern, highlighting its prevalence, can provide valuable guidance for the development of program repair tools.

\todo{this next paragraph doesn't seem to fit here, do we need it?}
A general approach for code embedding is to extract paths with semantic information that better serve a given embedding goal. For instance, to design an embedding method for repairing Rust programs, the designer might want to disregard certain elements, such as moving ownership to closure bodies. Such elements may be irrelevant for repair because they did not appear in the bug patterns that we presented.

@@

There are two main threats to the internal validity of our work: (1) A threat to internal validity is confounding, where changes to what shows up in the embedding are not due to changes in the code being embedded. In this case, the weighting scheme may contribute to confounding because we manually adjusted the weights. Nevertheless, our weighting scheme can be readjusted to find different patterns, which is why we made our pipeline publicly available. (2) Our code embedding approach is based on the frequency of observed program elements in ASTs, and we use DBSCAN as our clustering algorithm. Other code embedding methods and clustering algorithms (e.g. SLINK) might output new clusters that our pipeline is unable to find.

A threat to external validity is selection bias, where the benchmarks differ systematically from the actual world. The selected projects might cause selection bias and therefore we might not have presented patterns that may exist in other projects. Another threat to external validity can be the developers reporting the commits as bug fixing commits while they are not really fixing a functionality. Similarly, a commit message might not contain our target keywords while the commit is associated with bug fixing changes.

### 2023-05-27 results.tex
rework most useful patterns and add more todos

\textbf{What are the general fix patterns in Rust, and how often do they apply?}

We collected 60583 datapoints for general fix patterns using a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to compute both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters, resulting in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column). Since these patterns are the general Rust patterns, analogous patterns can also occur in other programming languages, and we will discuss such patterns when describing each pattern. We will also discuss the actionability of some of the patterns. Moreover, no matter whether we can find actionable outcomes for the patterns, other researchers can use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns. We next provide a detailed description of each of these patterns:

@@

Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute.

The concept of attributes in Rust is similar to Python's decorators and Java's annotations (though Java annotations may be less likely to have runtime effect; for instance, Java @Override adds additional compile-time constraints). While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. \\

@@

[discussion on how many of the changes are of each variant: over 90\% are modify]

@@

\todo{please confirm: G.attr.struct.field is not a subset of G.attr.struct.}

\noindent\textbf{Actionability:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Because so many of our bug fixes change attributes, we believe that IDEs ought to be able to propose relevant attribute changes, as witnessed in our corpus. The example in the text about changing Deserialize to ConfigDeserialize is specific to that program, but other changes would apply more generally. \todo{which?} Also, we demonstrated a case where the developer removed a \texttt{Debug} attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes. \\

@@

\noindent\textbf{Actionability:} When the user's specification requires a variable to be able to assume \texttt{None} values, these patterns indicate the need to modify the program accordingly. A program repair tool can leverage these patterns to automatically update the variable's type to \texttt{Option<T>}, ensuring that it can now hold \texttt{None} as a valid value. Furthermore, such a tool should also systematically update other occurrences of that variable throughout the program to account for the change and ensure program correctness.

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This concept might be referred to as interfaces in other languages, although there might be small differences. This change pattern relates to modifications in trait bounds. This can happen as part of bug fixes or for refactoring. \\

@@

In Rust, match control flow construct is used for comparing a value against multiple patterns, and executing the code associated with the first pattern that matches with our value. A pattern and its associated code are called a match arm. Match statements can be similar to switch statements in other languages, like C, although match statements in Rust are able to match more complicated patterns.\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. It is important to note that due to the unique nature of the Borrow-Checker as an exclusive component of Rust, these patterns may not have direct counterparts or equivalents in other languages. Now we provide a detailed description of each of these patterns:

@@

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. Since detecting whether a clone is redundant is undecidable, Clippy conservatively estimates redundancy. By definition, removing a clone that is indeed redundant does not modify functional aspects of the program behaviour. (Removing a non-redundant clone introduces a compile error or, worse yet, a bug.) However, such a removal can greatly improve performance, if the program was formerly cloning large objects. \\

@@

\noindent\textbf{Actionability:} A program repair tool can implement clone removal by examining the necessity of the clone operation and considering the user's specifications regarding variable cloning. If the clone is unnecessary, the tool can replace it with a reference to reduce the overhead of unnecessary copying, as discussed in the example above. \\

@@

To give an example, the commit in starship/starship excerpted above\footnote{\scriptsize \url{https://github.com/starship/starship/commit/56d475578ea508631275772127f49a6949fea6b0}} shows that the developer now borrows the value of \verb+repo_dir+ to keep the ownership of the object in the current scope. This object can then be used in a subsequent function invocation (\verb+remove_dir_all(repo_dir)+), which fixes a bug. \\

\noindent\textbf{Actionability:} This pattern often occurs as part of a larger change, making it challenging to create a generalized repair tool that covers all possible scenarios. Typically, when a reference is passed instead of transferring ownership, there is subsequent usage of the variable after the function call. Although creating a comprehensive repair tool for such cases is complex, empirical results underscore the significance and prevalence of this fix, warranting further investigation. \\

@@

\noindent\textbf{Description:} This change happens when the developer needs to mutate a variable which was previously immutable in a scope---they thus need to add the \verb+mut+ keyword. We never observed this pattern in refactoring changes; it always accompanied a change in program behaviour. \\

\noindent\textbf{Actionability:} A repair tool can utilize this pattern to meet the user's specifications by introducing mutability to the target variable and modifying other parts of the program accordingly. For instance, if the target variable has associated \texttt{impl} blocks, the change may involve modifying method annotations to ensure that the calling object is mutable, aligning with the desired specifications. \\

>>>>

\textbf{What are the general fix patterns in Rust, and how often do they apply?} 
We collected general fixes, asked DBSCAN to create clusters, and manually analyzed them. In this subsection, we will discuss each pattern, as well as its actionability---possible usages of the pattern for bug repair. Section~\ref{subsec:patterns} contains a global discussion of our patterns and their value.

Specifically, our dataset includes 60583 datapoints for general fix patterns. We processed our dataset on a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz). The data collection took around 26 hours to populate the general and BC-related databases $D_g$ and $D_b$. We used DBSCAN with parameters $\epsilon=0.0001$ and $Z=5$, resulting in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column).

Since these patterns are general Rust patterns, analogous patterns can also occur in other programming languages, and we will discuss such analogous patterns when describing each pattern. We will also discuss the actionability of some of the patterns. Moreover, other researchers can use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns; they are not limited to our suggested actionability. We next provide a detailed description of each of these patterns.

@@

Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes\todo{how important is inner versus outer?}: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute.

The concept of attributes in Rust is similar to Python's decorators\todo{are python decorator changes important in the Python dataset?} and Java's annotations (though Java annotations may be less likely to have runtime effect; for instance, Java @Override adds additional compile-time constraints). While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. \\

@@

\todo{discussion on how many of the changes are of each variant: over 90\% are modify, but get the actual number}

@@

\todo{please confirm: G.attr.struct.field is not a subset of G.attr.struct.}
\todo{we should also get the count of subtypes}

\noindent\textbf{Actionability:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Starting from the observation that so many of our bug fixes change attributes, we believe that IDEs need to be able to propose relevant attribute changes, as witnessed in our corpus. The example in the text about changing Deserialize to ConfigDeserialize is specific to that program, but other changes would apply more generally. \todo{which?} Also, we demonstrated a case where the developer removed a \texttt{Debug} attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes, where the developer needed to access a field which had not been marked for public access. On the other hand, public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes. \\

@@

\noindent\textbf{Actionability:} When the user's specification requires a variable to be able to assume \texttt{None} values, these patterns indicate the need to modify the program accordingly. A program repair\todo{is this really a repair tool or a refactoring tool? I feel like many of our changes are for refactorings not repairs, despite our PR} tool can leverage these patterns to automatically update the variable's type to \texttt{Option<T>}, ensuring that it can now hold \texttt{None} as a valid value. Furthermore, such a tool should also systematically update other occurrences of that variable throughout the program to account for the change and ensure program correctness.

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This concept more-or-less corresponds to interfaces in other languages, although the mapping is not exact (e.g. it is possible to implement traits for types that one did not initially define). This change pattern relates to modifications in trait bounds, which can happen as part of bug fixes or for refactoring. \\

@@

In Rust, the match control flow construct is used for comparing a value against multiple patterns, and executing the code associated with the first pattern that matches with our value. A pattern and its associated code are called a match arm. Match statements can be similar to switch statements in other languages, like C, although match statements in Rust are able to match more complicated patterns.\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We applied DBSCAN with parameters $\epsilon=0.001$ and $Z=5$, resulting in 102 clusters. After manual analysis (Section~\ref{sec:manual_analysis_cluster_selection}), we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. Since the Borrow-Checker is unique to Rust, patterns in this section generally do not have direct counterparts in other languages. We next provide a detailed description of each of these patterns.

@@

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. Since detecting whether a clone is redundant is undecidable, Clippy conservatively estimates redundancy. By definition, removing a clone that is indeed redundant does not modify functional aspects of the program behaviour. (Removing a non-redundant clone introduces a compile error or, worse yet, a bug.) However, such a removal can greatly improve the nonfunctional property of performance, if the program was formerly cloning large objects. \\

@@

\noindent\textbf{Actionability:} Like Clippy, a program repair tool can implement clone removal by reasoning about whether the clone operation was necessary. If the clone is unnecessary, the tool can replace it with a reference (as seen above), thus reducing the overhead of unnecessary copying. \\

@@

As an example, the commit in starship/starship excerpted above\footnote{\scriptsize \url{https://github.com/starship/starship/commit/56d475578ea508631275772127f49a6949fea6b0}} shows that the developer now borrows the value of \verb+repo_dir+ to keep the ownership of the object in the current scope. This object can then be used in a subsequent function invocation (\verb+remove_dir_all(repo_dir)+), which fixes a bug. \\

\noindent\textbf{Actionability:} This pattern often occurs as part of a larger change, making it challenging to create a generalized repair tool that covers all possible scenarios. Typically, when a reference is passed instead of transferring ownership, the variable is used again after returning from the call. Although creating a comprehensive repair tool for such cases is complex, empirical results underscore the significance and prevalence of this fix, warranting further investigation. \\

@@

\noindent\textbf{Description:} This change happens when the developer needs to mutate a variable which was previously immutable in a scope---they thus need to add the \verb+mut+ keyword. We never observed this pattern in refactoring-only changes; it always accompanied a change in program behaviour. \\

\noindent\textbf{Actionability:} A repair tool can use this pattern to introduce mutability to the target variable and modify other parts of the program accordingly. For instance, if the target variable has associated \texttt{impl} blocks, the change may involve modifying method annotations to ensure that the calling object is mutable, aligning with the desired specifications.\todo{how would the specs be provided? how could this be done as part of repair vs as part of refactoring?!} \\
